ggalluvial: Layered Grammar for Alluvial Plots

### 2020-03-16 paper.rmd
proofread joss paper

Figure \ref{fig:example} illustrates these and other plot elements by visualizing changes in several students' curricula (based on their declared majors) across several academic terms.
Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc. The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but switched to Painting in the 11th term, where the ribbon turns pink. The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves. When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.

@@

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, it is useful to adopt a careful demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible.
The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
Software implementations may use heuristic algorithms to position graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
Arguably, the extensions are better understood as producing a different type of diagram.

@@

Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer one-parameter families that interpolate between straight and x-spline ribbons.[^curves]
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology introduced above.
These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded. The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that I believe the new typology and terminology are warranted.

@@

**Repeated ordinal measures data:**
Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination. Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well. A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order. In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.

**Incomplete longitudinal data:**
Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification. While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the dimensional axis.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.

**Signed categorical data:**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign. Future use cases may demonstrate additional practical uses for this functionality.

>>>>

Figure \ref{fig:example} illustrates these and other plot elements by visualizing changes in several students' curricula (based on their declared majors) across several academic terms.
Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc. The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but had switched to Painting by the 11th term, where the ribbon turns pink. The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves. When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.

@@

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, I have found it useful to adopt a careful demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible.
The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
Software implementations may use heuristic algorithms to position their graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
Arguably, these extensions are better understood as producing a different type of diagram.

@@

Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing, and optionally weighted).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer one-parameter families that interpolate between straight and x-spline ribbons.[^curves]
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over the statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces, with the geological terminology introduced above.[^plots]
These alluvial plots are distinguished by two features: a prescribed order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
In combination, these features confer greater meaning on the second plotting dimension.

The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could, for example, arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
These features are particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that I believe the new typology and terminology are warranted.

@@

**Repeated ordinal measures data.**
Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size (though some follow user-provided hierarchies), and most insert gaps between categories for easy visual discrimination. Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories (hence of gaps) changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well. A use case by @Schlotter2019, to represent patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order. In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.

**Incomplete longitudinal data.**
Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification. While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the main dimension.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.

**Signed categorical data.**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even when cases change sign. Future applications may demonstrate additional uses for this functionality.
